\documentclass{article}

% conventions -
% Interfaces are in italics
% Module names in Bold
% method names in bold italics
% when describing the bioperl objects, list all
% the applicable interfaces and objects at the beginning
% of the subsection like
% Objects: Object1 Object2
% Interfaces: Interface1 Interface2

\usepackage {
 fullpage,fancyheadings,lscape,rotating,
 float,longtable,epsfig,psfig,layouts
}

\begin{document}

\begin{twocolumn}
\title{The Bioperl Project}
% Not sure how to do this - some people should be in the title?
\author{The BioPerl Project, http://bio.perl.org}
\maketitle

\begin{abstract}
The Bioperl project is an open-source collaboration of biologists,
computation biologist, bioinformaticians, and computer programmers.
The mission of the project is to provide an accessible programming
toolkit for manipulating and managing life science information.
\end{abstract}

\section{Introduction and History}

In this paper we'll describe the history of the Bioperl project, the
initial goals and how we've met them, examples of Bioperl modules at
work in other projects, and future directions for the project.

In 1995 there were few programmimg toolkits for manipulating
biological data and results from sequence analysis and there were
certainly none in Perl.  The GCG package \cite{GCG} was used for
analyzing sequence data and determining phylogenetic trees.  Users
could automate analysis by writing scripts to pass files between GCG
programs.  However, a specific file format was required and
integration of results from different sequence analysis packages was
difficult to do computationally. Other tools such as the NCBI toolkit,
were being developed to provide applications for running sequence
comparision and database searching on a biologist's local machine.
The difficulty with all these packages is they required and intimate
knowledge of the command line arguments and which parameters to change
for desired.  Additionally the output from all the analysis packages
represented similar information but in different formats.  A user
attempting to compare results across environments found it difficult
to produce a simple summary table because of the sophisticated parsing
script he might have to write. \\

The Bioperl project grew out of a need to manipulate biological data
in many different file formats and sources.  Perl was an ideal
language because it can be used to write quick throw away scripts as
well as reusable code libraries.  While Perl does have limitations, it
provides an easy platform for writing simple bioinformatics solutions
for neophyte computer programming biologists and advanced
computational biologists alike.

\section{Project Goals} 

The goals of the Bioperl project are to provide simple interfaces for
manipulating life science data.  The project attempts to provide these
interfaces as compact perl library modules which are  

\subsection{Early History}

The origin of the bioperl project can be traced back to the Fall of
1995 in which a small group of people primarily made up of Steven
Brenner, Georg Fullen, Richard Resnick, and XXX began a mailing list
\cite{oldbioperlmaillist} to discuss building biological modules 
In the Fall of 1995, Steve Brenner along with other graduates (Georg
Fullen) of the University of Bielefeld course in bioinformatics
decided to build a set of perl modules that would help them unify a
representation of a biological sequence in all their programs.  They
wrote simple reusable operations for this object such as reverse
complement or protein translation of an exon sequence.  Additional
modules were begun to handle parsing different sequence file formats
to read sequences in from different sources and additionally write
them out again in various sequence file formats.  Other modules began
to accumulate after Steve and Georg contributing their own modules and
implementations of solutions to common bioinformatic problems.

\subsection{Modern History}
% Chris D: need Help here  -jason
In the XX of 199X money was obtained from ?? to buy an alpha linux
server for bioperl source code repository and mail/ftp/web server space.
Through the hard work of Chris Dagdigian - running out of his home for
X months before the Genetics Institute donated server room space and
bandwith to the project.  Since then we have grown to 5 projects
including the BioJava and BioPython 'language' projects and BioXML and
BioCORBA standards projects.  These will be discussed later in this
paper.  
Formation of OpenBio...Hardware grant donation\\
\\ % Is this even relevant. -jason
The project was first coordinated in a 'benevolent dictator' style and
such great leaders as Steve Chervitz and Ewan Birney.  Later the
leadership was expanded to a core group of 4 developers and a sysadmin
(Ewan with Chris Dadigian, Hilmar Lapp, Heikki Lehvaslaiho and Jason
Stajich)


\section{Development Process}

The open-source model has lent itself to be extremely complementary to
the development process for a bioinformatics toolkit.  As
bioinformatics is very much a collaborative science, open source
development has encouraged a synergy of ideas.  With freely available
source, users of the toolkit can easily trace what modules do and how
to extend them for specialized cases.  Since anyone can contribute
their changes back to the toolkit other developers can take advantage
of their ideas and help maintain the code changes. 

\section{Design}

% Jason needs to rework!! --jason
Bioperl is an attempt to provide a fully Object-Oriented (OO) library of
perl modules for retrieving, storing, and manipulating life science
data especially in the sequence analysis corner of bioinformatics.
Attempts to make the toolkit fully OO have required the objects to be
defined as interfaces.  Since perl does not explictly have a
methodology for this we require all interface definitions in Bioperl
to be of the form ObjectI where 'I' indicates interface.  This can be
confusing when one tries to define a database interface without
indicating a subclass of the DBI module.  This document will italicize
all interface names and boldface all module names to help make clear
the distinction to the reader.  A preeceeding Bio:: indicating the
package namespace of all the Bioperl objects should be assumed for all
objects and interfaces discussed within this document.  
  
\par

We chose an OO design for the bioperl toolkit because it allows us to
properly assign tasks to objects and segregate functionality into
specialized objects rather than tackling all problems at once.
Utilizing interfaces one can separate the object's design from its
implementation.  This is particularly useful if one needs multiple
implementations of the same basic funcationality.  A good example of
this is the Sequence object in Bioperl. The \i{SeqI} interface
defines sequence objects which have associated sequence features.  It
is implemented by the \b{Seq} and \b{Seq::RichSeq} implementations
which serve as a simple and decorated implementation of the underlying
sequence.  Additionally the Ensembl project's VirtualContig object
implements the \i{SeqI} interface and represents a sequence
assemblies of a genome.  By implementing this in an Object-Oriented
bioperl methods can operate on the \b{Seq}, \b{Seq::RichSeq}, and
\b{EnsEMBL::Virtual::Contig} object in a uniform manner without
knowledge of which implementation is being processed.

\subsection{Sequence Objects}
Interfaces: \i{PrimarySeqI}, \i{SeqI}, \i{Seq::RichSeqI}
\\
Objects: \b{PrimarySeq}, \b{Seq}, \b{Seq::RichSeq}

Bioperl defines 2 types of Sequence objects, a \i{PrimarySeqI} and \i{SeqI}.
The \i{PrimarySeqI} (implemented by the class \b{PrimarySeq}) is a basic
sequence object containing a the DNA/RNA/Protein sequence as a perl
string.  The object also defines basic methods like 
\b{\i{length()}} - length of the sequence, 
\b{\i{moltype()}} - molecule type, 
\b{\i{recom}} - reverse complement of a DNA dnastring, 
\b{\i{trunc()}} - a new PrimarySeq object truncated which is 
a specified subsequence of the parent sequence, 
\b{\i{substr()}} - similar to trunc() except it only returns the sequence string rather than an
entire PrimarySeq object.  
\par 
The SeqI inherits from PrimarySeqI and is implemented by the Seq
object.  The object defines methods for storing, retrieving, and
removing sequence features (SeqFeatureI) and annotations (Annotation).

\subsection{Sequence Features and Annotations}

Interfaces: SeqFeatureI, AnnotationI
Objects: SeqFeature::Generic, SeqFeature

Sequence Features are conceptualized annotations with locations on a
sequence.  Sequence feature can be attached to a sequence via the
add_SeqFeature method in Bio::SeqI 

\subsubsection{Sequence Feature Locations}
Interfaces: LocationI, Location::SplitLocationI,
Location::FuzzyLocationI, Location::CoordinatePolicyI
\\
Objects: Location::Simple, Location::Split, Location::Fuzzy,
Location::NarrowestCoordPolicy, Location::WidestCoordPolicy,
Location::AvWithinCoordPolicy
\\
Sequence feature locations are a difficult entity to model.  Start and
stop points of an annotation can be unknown, a feature can be located
as a collection of small pieces of non-contiguous locations, or a
feature could be referenced only relative to another feature's
location.  These are difficult concepts to model in software which is
oriented towards absolute values rather than fuzzy ranges.   The
Location packages attempt to localize this into a set of classes which
can handle simple (start/end/strand), fuzzy (unclear or relative
start/end points), and split( a collection of non-contiguous fuzzy or simple
locations).
  

\subsection{Sequence File Parsing}

The SeqIO package was written to provide flexible parsing of sequence
files.  Supported formats include Fasta, GenBank, EMBL, SwissProt,
GCC, AceDB, PIR, Raw sequence, SCF, and GAME XML. 
% nice to get PIR back in here when the module works again...

The objects are accessed by creating a new SeqIO object and iterating
through available sequences via the next\_seq() method.  A simple
demonstration of using the object to convert between sequence formats
can be found in Figure XX.



\section{Sequence Databases}

Interfaces: DB::SeqI, DB::RandomAccessI

\subsection{Local Sequence Databases}
Objects: Index::Abstract, Index::AbstractSeq, Index::Fasta,
Index::EMBL, Index::GenBank, Index::SwissPfam, Index::Swissprot

\subsection{Remote Sequence Databases}

Objects: DB::GenBank, DB::GenPept, DB::EMBL, DB::SwissProt

\subsection{Parsing Analysis Results}

\subsubsection{Sequence Database Searching and Report Parsing}

Objects: Tools::Blast, Tools::BPlite, Tools::Fasta,
Tools::HMMER
Interfaces: SeqAnalysisParserI, Factory::SeqAnalysisParserFactoryI

Parsing result files from Sequence database searches is a common need
in bioinformatics.  Typically BLAST and FASTA are used to generate
thousands of reports of which a researcher may only want to see a
small fraction.  Parsing these reports into queriable objects allows
programmers to write software to select only the reports that meet a
user's criteria.
\\
The Bio::SeqAnalysisParserI is a generic interface that defines one
method - \bold{\i{next\_feature}}.  This method returns a Bio::SeqFeatureI object
and allows a programmer to write software which iterates through an
analysis report which may be from Blast (NCBI or WU-Blast), Bl2seq,
PsiBlast, HMMER, or Fasta. 

\subsubsection{Multiple Sequence Alignment}

% Heikki to help write

\section{Sequence Variation Toolkit}

\subsection{Sequence Variation Toolkit}

\subsection{LiveSeq Toolkit} 
\section{Use Cases}

We believe the Bioperl toolkit is useful in a variety of settings,
both in small and large scale projects.  The following examples
show that Bioperl is powerful enough to be used in production
environments while suitably lightweight for use in individual research
labs.

\subsection{Ensembl}
Project Ensembl is a automated whole genome annotation system
coordinated as a joint project between the Eurpoean Bioinformatics
Institute (EBI) and the Sanger Centre in Hinxton, UK.    


\section{Related projects}

Since Bioperl's inception a number of different open-source toolkit
projects have been started.  

\subsection{BioPython}

\subsection{BioJava}

\subsection{BioCORBA}

\subsection{BioXML}

\section{Contributors}

The primary contributors of the Bioperl project include the following
people along with others listed at http://bio.perl.org.

%-- see the AUTHORS file --%

\end{twocolumn}

\end{document}
